---
title: "Getting started with hydrographr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with hydrographr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
#bibliography: literature.bib
link-citations: yes
linkcolor: blue
#csl: copernicus.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

One of the major advantages of the package is that it can handle large geodata efficiently using a suite of Open Source geoprocessing tools like  `GRASS GIS`, `GDAL`, and `pktools`. With the`hydrographr` package large geodata is not loaded into R. Instead the files get processed in the background using `GRASS GIS`, `GDAL`, and `pktools` and only the output gets read into R. 

<!--Processing of large files often makes server inevitable and on servers Linux is the most common operating system. Therefore, `hydrographr` is optimized for Linux operating systems. However, we also implemented a way to use it within Rstudio on Windows. --> 

Since the `hydrographr` package is optimized to run in a Linux operating system, in a Windows environment some additional software needs to be installed (see [Windows system setup](https://glowabio.github.io/hydrographr/articles/windows_system_setup.html)).

## System requirements

To work smoothly with the `hydrographr` package, `GRASS GIS`, `GDAL`, and `pktools` need to be installed. 

Here you can find the installation guideline for your operating systems:

[Linux](https://glowabio.github.io/hydrographr/articles/linux_system_setup.html)

[Windows](https://glowabio.github.io/hydrographr/articles/windows_system_setup.html)

[macOS](https://glowabio.github.io/hydrographr/articles/macos_system_setup.html)


## Loading `hydrographr`

You can install `hydrographr` from its GitHub repository. If you did not install the `R` package yet, install it with `remotes::install_github()`.

```{r, eval=FALSE}
# If the package remotes is not installed run first:
install.packages("remotes")

remotes::install_github("glowabio/hydrographr")
```

Before we start exploring the package load `hydrographr`.

```{r}
library(hydrographr)
```


<!--## `hydrographr` tutorial

In the how-to-get-started tutorial we will show ...

The the test dataset of `hydrographr` includes ..... -->
